---
title: "Maize Yield Uniformity Trial Analysis"
author: "sriniva sagathevan"
date: "2026-09-10"
output: html_document
---

### INTRODUCTION
This project uses a maize uniformity trial dataset (Saunders) to answer two questions:
1) was average yield higher in 1930 than in 1929, and does yield vary across field row
2) position, with a similar pattern in both years?

## data viewing

```{r}
library(agridat)
library(tidyverse) 
library(ggplot2) 
```

## data understanding

```{r}
data("saunders.maize.uniformity")
nrow(saunders.maize.uniformity)
ncol(saunders.maize.uniformity)
head(saunders.maize.uniformity)
?saunders.maize.uniformity
summary(saunders.maize.uniformity)
```

## data cleaning

```{r}
sum(is.na(saunders.maize.uniformity$row))
sum(is.na(saunders.maize.uniformity$col))
sum(is.na(saunders.maize.uniformity$year))
sum(is.na(saunders.maize.uniformity$yield))
boxplot(saunders.maize.uniformity$yield)$out
```

The dataset had 17 missing yield values and one outlier (19.9), both removed before analysis.

```{r}
saunder_cleaned <- saunders.maize.uniformity %>%
  select(row,col,yield,year) %>%
  filter(complete.cases(.)) %>%
  filter(yield !=19.9)

nrow(saunder_cleaned)
boxplot(saunder_cleaned)$out
```


### data manipulation
Analysis: Yield by Year

```{r}
saunder_cleaned <- saunders.maize.uniformity %>%
  select(row,col,yield,year) %>%
  filter(complete.cases(.)) %>%
  filter(yield !=19.9) %>%
  group_by(year) %>%
  summarise(average_yield=mean(yield)) %>%
  arrange(desc(average_yield))

saunder_cleaned
```

## Visualization: Yield by Row Position

```{r}
saunder_cleaned2 <- saunders.maize.uniformity %>%
  select(row, col, yield, year) %>%
  filter(complete.cases(.)) %>%
  filter(yield != 19.9)

ggplot(saunder_cleaned2
       , aes(x = row, y = yield, color = factor(year))) +
  geom_point(size=1,alpha=1)+
labs(title="Maize yield by field row position")
```

### conclusion 
 this project answers two question 
1) was the average yield higher in 1930 than in 1929 
ans : the yield was increased from 1929 to 1930

2)Does yield vary across field row
ans : yes, it varied and follows a pattern in both the year 1929 and 1930.





